Dark matter halo shape in the Auriga simulations: radial dependence, time evolution and correlation with baryon disk properties

### 2019-07-17 Document/paper.tex
esquelto
edit

The DM halo shape at a fixed radius is an estimate of either
the isopotential or isodensity surfaces.  
Observational inference models usually estimate the 
isopotential contours which are probed by tracers (gas, stars), while
simulations work with the isodensity contours which can be directly
calculated from particle positions.  
Furthermore, the density contours in thin shells are very sensitive to
the presence of small satelites.  
For this reason we measure the shape by taking
volume-enclosed particles, rather than shell-enclosed.  
This method yields results in good agreement to the isodensity
contours in the inner halo region, roughly for radii smaller than one
quarter the virial radius as explored by
\citep{Vera-Ciro_et_al._2011}.

We measure the shape using the reduced inertia tensor \citep{Allgood_et_al._2006},

@@

We start the calculations taking into account particles within a
sphere of radius $R$ and then recharacterize the triaxial parameters
by taking into account particles within an ellipsoid of semi-axes
$r,r/q,r/s$ and re-scaled distance $d^2=x^2+(y/q)^2+(z/s)^2$, where $q
= b/a$ and $s=c/a$ are the previously calculated axial ratios. 
We repeat this process until the average deviation of semi-axes is
less than $10^{-6}$.  
This is the same method used to estimate the halo shape in the DM-only
Aquarius simulations \citep{Vera-Ciro_et_al._2011}.

Following the convergence critery by \cite{Vera-Crito_et_al.2011} we
restrict the sampling of the ellipsoidal parameters to radii  between
$\sim 2$kpc and $R_{200}$, where  $R_{200}$ correspond to the  radius
enclosing a sphere with 200 times the average dark matter density of
the Universe.  
On average, over the 30 halos in the \texttt{Level4} sample
$R_{200}=200\pm 10$kpc. 
For \texttt{Level3} halos we go down to distances $\sim 0.2$ kpc.

@@

In the DMO sample we find that halos are rounder with increasing
radius.
The upper panels in Figure \ref{fig:slices} illustrates this effect.
The contours show a projected DM slice while the ellipsoid corresponds
to the full 3D shape determination. 
There we see a highly ellipsoidal halo shape at radii $\sim 3$kpc
that becomes less triaxial at $\sim 50$ kpc.

We summarize this trend in Figure \ref{fig:triaxiality_plane} by
plotting the results of all the 30 halos in the DMO sample.
The left panel shows every halo in the $c/a$-$b/a$ plane at
two different radii $R_{200}/8 (\sim 20$kpc$)$ and $R_{200}$. 
The outer part of the halo is systematically rounder tha its inner
region. 
Nevertheless the halo shape can still be considerated to be prolate at
all radii. 
These plots confirm the results already reported in the
literature \citep{Vera-Ciro_et_al._2011}.

@@

Next we want to test to what extend the global effect of decreasing
triaxility in MHD simulations compared to the DMO sample 
holds for individual halos. 
We compute $\Delta T\equiv T_{\rm MHD}-T_{\rm DMO}$ the difference between the
triaxility in the MHD and the DMO simulation for each individual halo. 
Figure \label{fig:delta_triaxial_cumulative} shows the cumulative
distribution at the same radii as in
Figure \label{fig:delta_triaxial_cumulative}.

@@

In the previous section we stablished that DMO halos at $z=0$ have
a steady and monotenouse trend towards decreasing triaxility with
increasing radius.
\cite{Vera-Ciro_et_al._2011} found that this  radial trend at $z=0$ 
correlates with the halo shape as a function of cosmic time, i.e.
the halo becomes less triaxial as cosmic time progresses. 
This correlation was interpreted as each radial shell retaining memory
of the accreting conditions at the virial radius at the time of
collapse.

Here we use the six \texttt{level3} halos to measure to what extent
the presence of baryons affect this memory effect.
The left column of Figure \ref{fig:redshift_triaxial} shows the six
halos in the triaxility plane. 
The colour dots show the time evolution while the black line
corresponds to the radial trend at $z=0$.
The close overlap of the spatial and temporal trends confirms the
memory effect found by \cite{Vera-Ciro_et_al.2011}.

In the MHD sample the memory effect is less evident. 
The right column of Figure \ref{fig:redshif_triaxial} shows the
corresponding results for the six halos in the MHD sample.

%Discuss if this correlation may be recovered if compared for example
%at Disk radius in stead of virial radius.\\

@@

\subsection{What drives the rounding effect?}

@@

\section{Conclusions}

>>>>

The DM halo shape at a fixed radius is an estimate of either
the isopotential or isodensity surfaces.  
Observational inference models usually estimate the 
isopotential contours which are probed by tracers (gas, stars), while
simulations work with the isodensity contours which can be directly
calculated from particle positions.  
Furthermore, the density contours in thin shells are very sensitive to
the presence of small satelites.  
For this reason we measure the shape by taking
volume-enclosed particles, rather than shell-enclosed.  
This method yields results in good agreement to the isodensity
contours for radii $leq 140$ kpc as explored by
\citep{VeraCiro11}.

In particular, we measure the shape using the reduced inertia tensor
\citep{Allgood_et_al._2006},

@@

We start the calculations taking into account particles within a
sphere of radius $R$ and then recharacterize the triaxial parameters
by taking into account particles within an ellipsoid of semi-axes
$r,r/q,r/s$ and re-scaled distance $d^2=x^2+(y/q)^2+(z/s)^2$, where $q
= b/a$ and $s=c/a$ are the previously calculated axial ratios. 
We repeat this process until the average deviation of semi-axes is
less than $10^{-6}$.  
After convergence we define a unique radius $R$ as the geometrical
mean of the axial lenghts $R=(abc)^{1/3}$.
We use this radial coordinate $R$ to parameterize the spatial changes
in halo shape we report in the following sections.

This is the same method used to estimate the halo shape in the DM-only
Aquarius simulations \citep{VeraCiro11}.

Following the convergence criterion by \cite{Vera-Crito_et_al.2011} we
restrict the sampling of the ellipsoidal parameters to radii  between
$\sim 2$kpc and $R_{200}$, where  $R_{200}$ correspond to the  radius
enclosing a sphere with 200 times the critical density of the Universe.
On average, over the 30 halos in the \texttt{Level4} sample
$R_{200}=230\pm 15$kpc. 
For \texttt{Level3} halos we go down to distances of $\sim 0.2$ kpc.

@@

In the DMO sample we find that halos are rounder with increasing
radius.
The upper panels in Figure \ref{fig:slices} illustrate this effect.
The contours show a projected DM slice while the ellipsoid corresponds
to the full 3D shape determination. 
There we see a highly ellipsoidal halo shape at radii $\sim 3$kpc
that becomes less triaxial at $\sim 50$ kpc.

We summarize this trend in Figure \ref{fig:triaxiality_plane} by
plotting the results of all the 30 halos in the DMO sample.
The left panel shows every halo in the $c/a$-$b/a$ plane at
two different radii $R_{200}/8 (\sim 20$kpc$)$ and $R_{200}$. 
The outer part of the halo is systematically rounder than its inner
region. 
Nevertheless the halo shape can still be considerated to be prolate at
all radii. 
These plots confirm the results already reported in the
literature \citep{VeraCiro11}.

@@

To quantify to what extend the global effect of decreasing
triaxility in MHD simulations compared to the DMO sample 
holds for individual halos. 
We compute $\Delta T\equiv T_{\rm MHD}-T_{\rm DMO}$ the difference between the
triaxility in the MHD and the DMO simulation for each individual halo. 
Figure \label{fig:delta_triaxial_cumulative} shows the cumulative
distribution at the same radii as in
Figure \label{fig:delta_triaxial_cumulative}.

@@

In the previous section we stablished that DMO halos at $z=0$ have
a steady and monotonous trend towards decreasing triaxility with
increasing radius.
\cite{VeraCiro11} found that this  radial trend at $z=0$ 
correlates with the halo shape as a function of cosmic time, i.e.
the halo becomes less triaxial as cosmic time progresses. 
The hypotesis put forward by \cite{VeraCiro11} is that the correlation
raises as each radial shell retains memory of the accreting conditions
at the virial radius at the time of collapse.

Here we use the six \texttt{level3} halos to measure to what extent
the presence of baryons affect this memory effect.
In Figures \ref{fig:triaxial_history_A} and
\ref{fig:triaxial_history_B} we follow the shape evolution measured at
$R_{200}$ since redshift $z=1$ to the present, these results are
represened by the coloroured dots.
We compare this temporal evolution against the radial evolution at
$z=0$ represented by the dash black lines.

The left column of Figure \ref{fig:redshift_triaxial} shows the six
halos in the triaxility plane. 
The colour dots show the time evolution while the black line
corresponds to the radial trend at $z=0$.
The left/right columns correspond to the DMO/MHD simulations, respectively.

We start by analyzing to what extent the memory effect is
already present in the DMO sample. 
The first feature that jumps out is that \texttt{Au-21} in the DMO
simulation does not have a simple shape evolution history.
By visual inspection we find that an extended merging substructure
intereacted in the main halo showing up in the shape measurements,
making it more difficult to interpret the memory hypothesis.
For this reason we exclude this halo from the analysis.

For the other five halos we find simpler temporal and spatial
evolution histories. 
Halo \texttt{Au-6} (Figure
\ref{fig:triaxial_history_A}) shows a broad
correlation between the two tracks; close to the cases of halos
\texttt{Aq-4-C} and \texttt{Aq-4-D} in \cite{VeraCiro11}.
The four other halos (\texttt{Au-16} in Figure
\ref{fig:triaxial_history_A}; \texttt{Au-23}, \texttt{Au-24},
\texttt{Au-27} in Figure \ref{fig:triaxial_history_B}) are four
cases where the temporal evolution if followed very closely by the
radial trends.
We conclude that in the DMO sample 4 out of the 5 high resolution
halos closely follow the memory hypothesis trends and the remaning
halo shows a reasonable correlation.

The situation in the MHD simulations is more diverse.
As expected by the results in the previous section the broad location
in the triaxility plane of the temporal and radial evolution in the
MHD simulation is completeley different from its DMO counterpart. 
In terms of the memory hypothesis we see two different trends.
We have the case of Halos \texttt{Au-6} and \texttt{Au-16}

@@

\subsection{Correlation with baryonic disk properties}

@@

{\bf Comparison with other detailed simulations.}

{\bf Comparison with other statistical simulations.}

{\bf Comparison against observations.}

{\bf Implications for observational studies in our Galaxy}

\section{Discussion and Conclusions}

In this paper we measured the shape of thirty isolated Milky Way like 
dark matter haloes simulated in the Auriga project using the zoom-in
technique. 
Six of these haloes are also simulated at higher resolution.
The simulations
